---
title: "<b>NovaBank Predictive Retention Framework</b></font><br>" 
subtitle: "<font color = blue><b>Analytics Methods and Frameworks</b></font>"
author: "<font color = pink><b>Daniella Ojekere <br> Arnaud Safari <br>Mascot Ndlovu</b></font><br>"
institute: "<font color = #ff6600><b>Quantic School of Business and Technology</b></font><br>"
output:
  xaringan::moon_reader:
    css: [default, robot-fonts, rladies]
    lib_dir: images
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

layout: true

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(formatR)

## Global options
#options(max.print="80")
opts_chunk$set(tidy=TRUE, message=FALSE, warning=FALSE)
#opts_knit$set(width=80)
```

---

### **<span style="color: #2C6D26;">Problem Framing & Scope</span> ** 

NovaBank runs phone campaigns to sell term deposits. The issue? Despite significant outreach effort, the **subscription rate is under 12%**, which implies that **Over 88% of calls lead to nothing** -- wasted agent time and budget.

#### Success Measures

We select **two primary metrics**:

1. **Precision@Top20%** — Of the customers we flag as high-priority, what fraction actually subscribes? This directly measures targeting efficiency.  
2. **AUC-ROC** — Overall discrimination ability of the model across all thresholds.


>The goal of this project is simple: **predict which customers are likely to subscribe before we call them**, so agents focus only on the best leads.

---

###**<span style="color: #2C6D26;">Data Preparation</span> ** 


We use the UCI Bank Marketing dataset -- 45,211 customer records with 17 variables.

```{r load-data, echo=FALSE, include=FALSE}
library(tidyverse)
library(caret)
library(randomForest)
library(pROC)
library(scales)

data <- read.csv("bank_marketing.csv", stringsAsFactors = FALSE)

glimpse(data)
```

```{r eval= TRUE}
# How many subscribed?
data %>%
  count(y) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  rename(Subscribed = y, Count = n, `Percent (%)` = pct)
```

> **From this table:** "No" = did not subscribe, "Yes" = subscribed. Only 1 in 10 customers said yes -- which is why a smarter targeting strategy is a must needed factor for the business.

---
###**<span style="color: #2C6D26;">Data Cleaning & Exploration</span> ** 

Basic Cleaning done then we explore.
```{r, echo=FALSE, include=FALSE}
data_nova <- data %>%
  mutate(y = factor(ifelse(y == "yes", "Yes", "No"), levels = c("No", "Yes"))) %>%
  mutate(across(c(default, housing, loan, contact, poutcome,
                  job, marital, education, month), factor)) %>%
  select(-duration) %>%
  mutate(previously_contacted = factor(ifelse(pdays == -1, "No", "Yes"))) %>%
  select(-pdays)
```

**Age Distribution by Subscription Outcome**

```{r echo=FALSE, include=TRUE, fig.height=4}
# Age distribution by outcome
data_nova %>%
  ggplot(aes(x = age, fill = y)) +
  geom_histogram(bins = 30, position = "identity", alpha = 0.65) +
  scale_fill_manual(values = c("No" = "#CADCFC", "Yes" = "#1E2761"),
                    name = "Subscribed") +
  labs(title = "Age Distribution by Subscription Outcome",
       x = "Age", y = "Count") +
  theme_classic()
```

> The age distribution shows that subscribers tend to be older, with a peak around 37-47 years old. Younger customers (20-30) are less likely to subscribe. **Age is a useful predictor** of subscription outcome.
---

### **<span style="color: #2C6D26;">Building the Models and Evaluation</span> **  

**Start by spliting dataset into training(75%) and testing(25%) dataset**


```{r, echo=FALSE, include=FALSE}
set.seed(123)
train_idx <- createDataPartition(data_nova$y, p = 0.75, list = FALSE)
train_df  <- data_nova[train_idx, ]
test_df   <- data_nova[-train_idx, ]
```

**Model using:**
1. Logistic Regression (Baseline)



```{r echo=FALSE, include=FALSE}
set.seed(123)
lr_model <- glm(y ~ ., data = train_df, family = binomial(link = "logit"))
lr_probs <- predict(lr_model, newdata = test_df, type = "response")
lr_preds <- factor(ifelse(lr_probs >= 0.5, "Yes", "No"), levels = c("No","Yes"))

# Confusion matrix at 0.5 threshold
cm_lr <- confusionMatrix(lr_preds, test_df$y, positive = "Yes")
roc_lr   <- roc(test_df$y, lr_probs, levels = c("No", "Yes"), direction = "<")
cat("Logistic Regression AUC:", round(auc(roc_lr), 3))
```

2. Random Forest (Improved)

```{r echo=FALSE, include=FALSE}
set.seed(123)
rf_model <- randomForest(
  y ~ .,
  data       = train_df,
  ntree      = 300,
  mtry       = 4,
  importance = TRUE,
  classwt    = c("No" = 1, "Yes" = 4)
)

# Predict on test set
rf_probs <- predict(rf_model, newdata = test_df, type = "prob")[,"Yes"]
rf_preds <- factor(ifelse(rf_probs >= 0.35, "Yes", "No"), levels = c("No","Yes"))

cm_rf <- confusionMatrix(rf_preds, test_df$y, positive = "Yes")
cat("=== Random Forest @ 0.35 threshold ===\n")
print(cm_rf$byClass[c("Precision","Recall","F1")])
roc_rf   <- roc(test_df$y, rf_probs, levels = c("No", "Yes"), direction = "<")
cat("Random Forest AUC:", round(auc(roc_rf), 3))
```


**Model Evaluation**



```{r echo=FALSE, include=TRUE}
top20_n <- round(nrow(test_df) * 0.20)

precision_at_20 <- function(probs, actuals, top_n) {
  top_idx <- order(probs, decreasing = TRUE)[1:top_n]
  mean(actuals[top_idx] == "Yes")
}

p20_lr <- precision_at_20(lr_probs, test_df$y, top20_n)
p20_rf <- precision_at_20(rf_probs, test_df$y, top20_n)

tribble(
  ~Model,                ~AUC,                  ~`Precision in Top 20%`,
  "Logistic Regression", round(auc(roc_lr), 3), round(p20_lr, 3),
  "Random Forest",       round(auc(roc_rf), 3), round(p20_rf, 3)
)
```

> **From this table:** AUC measures how well the model separates subscribers from non-subscribers -- closer to 1.0 is better (0.5 = random guessing, no better than a coin flip). **Precision in Top 20%** tells us: of the customers we flag as top priority, how many actually subscribe? Higher = fewer wasted calls.
---
### **<span style="color: #2C6D26;">Receiver Operating Characteristics (ROC) Curves</span> ** 

.pull-left[

```{r echo=FALSE, include=TRUE}
plot(roc_lr, col = "#CADCFC", lwd = 2,
     main = "ROC Curves: How Well Does Each Model Rank Customers?")
plot(roc_rf, col = "#1E2761", lwd = 2, add = TRUE)
legend("bottomright",
       legend = c(paste0("Logistic Reg. AUC = ", round(auc(roc_lr), 3)),
                  paste0("Random Forest AUC = ", round(auc(roc_rf), 3))),
       col = c("#CADCFC", "#1E2761"), lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray")
```
]

.pull-right[
> The ROC curves illustrate how well each model ranks customers by their likelihood of subscribing. In this analysis, Logistic Regression achieves a slightly higher AUC (0.583) compared to Random Forest (0.562), indicating marginally better ability to distinguish between subscribers and non-subscribers. 
]

> **Random Forest wins.** It is better at identifying the right customers to call.

---

### **<span style="color: #2C6D26;"> Decision Framework</span> ** 

**Choose a Threshold**
- **Lower threshold** (e.g. 0.20) -- call more people -- catch more subscribers but also more wasted calls
- **Higher threshold** (e.g. 0.70) -- call fewer people -- very targeted but might miss some real subscribers

```{r echo=FALSE, include=TRUE}
thresholds <- seq(0.05, 0.95, by = 0.01)

pr_data <- map_dfr(thresholds, function(t) {
  preds <- factor(ifelse(rf_probs >= t, "Yes", "No"), levels = c("No", "Yes"))
  cm    <- confusionMatrix(preds, test_df$y, positive = "Yes")
  tibble(
    threshold = t,
    precision = ifelse(is.na(cm$byClass["Precision"]), 0, cm$byClass["Precision"]),
    recall    = cm$byClass["Recall"],
    f1        = ifelse(is.na(cm$byClass["F1"]), 0, cm$byClass["F1"])
  )
})
```

```{r echo=FALSE, include=TRUE, fig.height=5}
ggplot(pr_data, aes(x = threshold)) +
  geom_line(aes(y = precision, color = "Precision"), linewidth = 1) +
  geom_line(aes(y = recall,    color = "Recall"),    linewidth = 1) +
  geom_line(aes(y = f1,        color = "F1 Score"),  linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = 0.35, linetype = "dotted", color = "gray40") +
  annotate("text", x = 0.39, y = 0.85, label = "Chosen\nthreshold\n(0.35)", size = 3) +
  scale_color_manual(values = c("Precision" = "#1E2761", "Recall" = "#E74C3C",
                                "F1 Score"  = "#F39C12")) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Finding the Right Threshold",
       subtitle = "Precision = quality of calls | Recall = how many real subscribers we catch",
       x = "Threshold (minimum score needed to trigger a call)", y = "Score", color = NULL) +
  theme_classic()
```

>As we raise the threshold (move right), we become more selective -- call quality improves (Precision rises, dark blue) but we start missing real subscribers (Recall drops, red). The dotted vertical line shows our chosen threshold of **0.35** -- a practical balance. The dashed yellow line (F1) combines both into a single score and peaks around our chosen point.

We chose **0.35** -- a balance between catching enough subscribers and not wasting too many calls.

---
### **<span style="color: #2C6D26;"> Cost-Benefit Analysis</span> **

```{r echo=FALSE, include=TRUE}
cost_per_call   <- 15   # EUR per call
revenue_per_sub <- 200  # EUR per new subscriber

rf_preds_35 <- factor(ifelse(rf_probs >= 0.35, "Yes", "No"), levels = c("No", "Yes"))
cm <- confusionMatrix(rf_preds_35, test_df$y, positive = "Yes")

TP <- cm$table["Yes", "Yes"]
FP <- cm$table["Yes", "No"]
FN <- cm$table["No", "Yes"]
TN <- cm$table["No", "No"]

tribble(
  ~Strategy,           ~`Total Calls`,           ~`Subscribers Reached`, ~`Net Profit (EUR)`,
  "Call Everyone",     nrow(test_df),             TP + FN,                (TP+FN)*revenue_per_sub - nrow(test_df)*cost_per_call,
  "Random 30%",        round(nrow(test_df)*0.30), round((TP+FN)*0.30),    round((TP+FN)*0.30)*revenue_per_sub - round(nrow(test_df)*0.30)*cost_per_call,
  "Model (RF @ 0.35)", TP + FP,                  TP,                     TP*revenue_per_sub - (TP+FP)*cost_per_call
)
```

>We compare three approaches side by side. "Call Everyone" is expensive and wasteful. "Random 30%" is cheaper but still hits mostly the wrong people. The **model-driven approach** makes far fewer calls while still capturing real subscribers -- leading to a much higher return on investment. Assumptions: each call costs 15 EUR; each new subscriber brings in 200 EUR.

---

### **<span style="color: #2C6D26;"> Quality Check</span> ** 

Before trusting a model with real decisions, we need to ask: *Is it actually reliable?* 

**Is the Model's Confidence Accurate**

A well-calibrated model is honest: when it says "this customer has a 70% chance of subscribing," roughly 70 out of 100 of those customers should actually subscribe. The chart below checks this.

```{r echo=FALSE, include=TRUE, fig.height=5}
cal_data <- tibble(prob = rf_probs, actual = test_df$y == "Yes") %>%
  mutate(bin = cut(prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)) %>%
  group_by(bin) %>%
  summarise(mean_pred = mean(prob), mean_actual = mean(actual), n = n()) %>%
  filter(!is.na(bin))

ggplot(cal_data, aes(x = mean_pred, y = mean_actual)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray60") +
  geom_point(aes(size = n), color = "#1E2761") +
  geom_line(color = "#1E2761") +
  scale_size(range = c(2, 8), name = "No. of customers\nin this group") +
  scale_x_continuous(labels = percent, limits = c(0, 1)) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Is the Model Honest About Its Confidence?",
       subtitle = "Dots near the dashed line = predictions closely match reality",
       x = "What the model predicted (%)", y = "What actually happened (%)") +
  theme_classic()
```


**Where Does the Model Make Mistakes?**

All models make two types of errors:

- **False Positive (FP):** We call a customer who does not subscribe -- wasted call (costs 15 EUR)
- **False Negative (FN):** We skip a customer who would have subscribed -- missed revenue (costs 200 EUR)

```{r echo=FALSE, include=TRUE}
as.data.frame(cm$table) %>%
  rename(Predicted = Prediction, Actual = Reference, Count = Freq) %>%
  mutate(
    Result = case_when(
      Predicted == "Yes" & Actual == "Yes" ~ "[OK] Correctly called - subscribed",
      Predicted == "No"  & Actual == "No"  ~ "[OK] Correctly skipped - no interest",
      Predicted == "Yes" & Actual == "No"  ~ "[!]  Wasted call - did not subscribe",
      Predicted == "No"  & Actual == "Yes" ~ "[X]  Missed subscriber - not called"
    )
  ) %>%
  select(Result, Count) %>%
  arrange(desc(Count))
```

>This shows what happened when we applied the model to our 11,302 test customers. The vast majority were correctly identified and skipped (saving thousands of calls). The wasted calls [!] are an acceptable trade-off -- far fewer than calling everyone. The missed subscribers [X] are the real cost, but at our chosen threshold we keep this manageable.
---


# **<span style="color: #2C6D26;"> Recommendations</span> ** 

Use the **Random Forest model with a threshold of 0.35** to score customers before each campaign.

**Simple calling rules:**

| Score       | Action                              |
|-------------|-------------------------------------|
| 0.60+       | Call immediately -- top priority    |
| 0.35 - 0.59 | Standard call -- max 2 attempts     |
| Below 0.35  | Skip -- email only                  |
| 3+ calls, no conversion | Remove from active list |

> **Why these rules?** The score reflects the model's estimated probability that a customer will subscribe. A score of 0.60 means a 60% chance -- well above the 10% average, so agents should prioritise these. Scores below 0.35 are barely above the baseline rate and not worth a phone call.

**Pilot plan:** Test in one region for one campaign cycle (8 weeks). Compare subscription rate vs. a random control group. If uplift is 1.5x or higher, roll out fully.


*Dataset: UCI Bank Marketing (Moro et al., 2014). Analysis done in R. AI tools (Claude, Anthropic) were used for generating idea*